Text embedding models enable high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-30 paper/main.tex
progress on results, split softwraps to hardwraps for shared editing

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can begin
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek \citep{Macl05, HallGree08}. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a
demonstration of how our models can be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy \citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work \citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model \citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model yielded a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics discovered by the model). Within this space, each lecture's sequence of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

@@

Next, we sought to characterize what information participants' performance on each of the three quizzes could provide about their conceptual knowledge at that point in time. Traditional approaches to evaluating students' performance on short-form knowledge assessments, such as those in our study, entail computing the proportion of correctly answered questions to assign a simple numeric score. While this may afford some measure of the \textit{extent} of a learner's knowledge, such a score provides little information to either the teacher or the learner about the particular \textit{contents} of their knowledge---in other words, raw ``proportion-correct'' measures may capture \textit{how much} a student knows, but not \textit{what} they know. For instance, suppose a participant in our study was highly knowledgeable about fundamental physical forces (i.e., lecture 1 content) but unfamiliar with how stars are formed (i.e., lecture 2 content), while a second participant was unfamiliar with fundamental forces but had extensive prior knowledge about star formation. Since quiz 1 (completed before viewing either lecture) contained an equal number of questions about each lecture's content (see Fig.~\ref{fig:experiment}), these two participants could easily achieve the same proportion-correct score despite their conceptual knowledge differing substantially. How might we distinguish these individuals?

We hypothesized that our text embedding model, which we had trained on transcripts of the two lectures to characterize their conceptual contents, should also allow us to capture the conceptual knowledge probed by each quiz question. If our model successfully represented information about the deeper conceptual content of the lectures (i.e., beyond surface-level details such as particular word choices) then we should be able to recover a correspondence between their embeddings and embeddings of other text that reflect related concepts (e.g., quiz questions that were ostensibly, by design, ``about'' one of the two lecture) despite differences in exact verbiage. Intuitively, while the content of each given lecture will naturally vary from moment to moment, the particular combination of dimensions along which its trajectory varies across moments correspond to the topics that encode meaningful ``information''~\citep{Fish22}  about that lecture's dynamic conceptual content. We therefore expected that quiz questions pertaining to those same concepts might express the same set of topics.

When we compared the variability in topic
weights across each lecture's timepoints with the variability in topic weights
across each question set, we found a strong correspondence
(Fig.~\ref{fig:topics}). The most variable topics from the \textit{Four
Fundamental Forces} lecture, and questions about that lecture, are 2, 3, 8, 12,
13, 14, and 15. The most variable topics from the \textit{Birth of Stars}
lecture, and questions about that lecture, are 1, 4, 5, 6, and 7. This strong
overlap between the lectures and questions specifically about each lecture
indicates that the topic model captures some of the underlying conceptual
content.

@@

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}
Following \citet{HeusEtal21}

>>>>

Our primary research goal is to advance our understanding of what it means to
acquire deep, real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can begin
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek~\citep{Macl05, HallGree08}. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a
demonstration of how our models can be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

We asked participants in our study to complete brief multiple-choice quizzes before, between, and after watching two lecture videos from the Khan Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form.
We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. The two videos we selected are introductory lectures that both
belong to Khan Academy's ``Cosmology and Astronomy'' course domain, but are taken from
different lecture series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies''
for the first and second lectures, respectively).

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work~\citep{HeusEtal21} to extract the latent themes from the lectures' contents
using a topic model~\citep{BleiEtal03}. Briefly, topic models take as input a
collection of text documents and learn a set of ``topics'' (i.e., latent
themes) from their contents. Once fit, a topic model can be used to transform
arbitrary (potentially new) documents into sets of ``topic proportions,''
describing the weighted blend of learned topics reflected in their texts. We
parsed automatically generated transcripts of the two lectures into
overlapping sliding windows, which we treated as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming these windows with our model yielded a number-of-windows by
number-of-topics (15) topic-proportions matrix, denoting the unique mixture
of broad themes from both lectures reflected in each window's content.
Intuitively, each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is analogous to a coordinate in a 15-dimensional
space (whose axes are topics discovered by the model). Within this space, each lecture's sequence of
topic vectors (i.e., corresponding to sliding windows parsed from its
transcript) forms a \textit{trajectory} that captures how its conceptual
content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We resampled
these trajectories to a resolution of 1 topic vector for each second of video.

Next, we sought to characterize what information participants' performance on
each of the three quizzes could provide about their conceptual knowledge at
that point in time. Traditional approaches to evaluating students'
performance on short-form knowledge assessments, such as those in our study,
entail computing the proportion of correctly answered questions to assign a
simple numeric score. While this may afford some measure of the
\textit{extent} of a learner's knowledge, such a score provides little
information to either the teacher or the learner about the particular
\textit{contents} of their knowledge---in other words, raw
``proportion-correct'' measures may capture \textit{how much} a student
knows, but not \textit{what} they know. For instance, suppose a participant
in our study was highly knowledgeable about fundamental physical forces (i.e
., lecture 1 content) but unfamiliar with how stars are formed (i.e., lecture
2 content), while a second participant was unfamiliar with fundamental forces
but had extensive prior knowledge about star formation. Since quiz 1
(completed before viewing either lecture) contained an equal number of
questions about each lecture's content (see Fig.~\ref{fig:experiment}), these
two participants could easily achieve the same proportion-correct score
despite their conceptual knowledge differing substantially. How might we
distinguish these individuals?

@@

We hypothesized that our text embedding model, which we had trained on
transcripts of the two lectures to characterize their conceptual contents,
should also allow us to capture the conceptual knowledge probed by each quiz
question. If our model successfully represented information about the deeper
conceptual content of the lectures (i.e., beyond surface-level details such
as particular word choices) then we should be able to recover a
correspondence between their embeddings and embeddings of other text that
reflect related concepts (e.g., quiz questions that were ostensibly, by
design, ``about'' one of the two lecture) despite differences in exact
verbiage. Intuitively, while the content of each given lecture will naturally
vary from moment to moment, the particular combination of dimensions along
which its trajectory varies across moments correspond to topics that encode
meaningful ``information''~\citep{Fish22}  about that lecture's dynamic
conceptual content. We therefore expected that quiz questions pertaining to
those same concepts might express the same set of topics.

%Figure~\ref{fig:topics} displays the variability in topic weights across timepoints in each lecture, and across questions in each category (i.e., lecture 1-related, lecture 2-related, and general physics knowledge).

We used our model to transformed the text of each question in our pool into
the same embedding space as the two lectures' trajectories (Fig
.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. When we compared the
variability in topic weights across each lecture's timepoints, and across
questions from each category (i.e., lecture 1-related, lecture 2-related, and
general physics knowledge), we found multiple interesting patterns. First,
there is a strong correspondence between the sets of topics expressed by each
lecture and its relevant questions. This indicates that our model is
sensitive to the underlying conceptual content tested by different question,
and should allow us to differentially assess participants' knowledge for the
two lectures based on the \textit{specific} questions they answered correctly
or incorrectly on each quiz. Second, the topics expressed by the general
physics knowledge questions comprise a nonspecific mix of topics from the two
lectures (4 from lecture 1; 5 from lecture 2), most of which are expressed to
a relatively low degree. This indicates that our model should enable us to
discriminate questions that are highly relevant to our assessment of
participants' knowledge about concepts from the lectures from those that are
not (or potentially less so). Third, the two lectures' trajectories vary
along largely non-overlapping topic dimensions (e.g., variations in weight
for topics 2, 3, 8, 12, 13, and 15 are specific to lecture 1; variations in
weight for topics 1, 4, 5, 6, 7, and 10 are specific to lecture 2), but also
share a small number of expressed topics in common (topics 11 and 14). This 
indicates that our lecture is able to simultaneously capture

@@

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}
Following~\citet{HeusEtal21}
